The Astropy Project: \\ Sustaining and Growing a Community-oriented Open-source Project and the Latest Major Release (v5.0) of the Core Package

### 2022-03-18 src/ms.tex
add git commit or tag name to latex header
starting to sketch challenges

\begin{abstract}
The \astropy Project supports and fosters the development of open-source and openly-developed
\python packages that provide commonly-needed functionality to the astronomical
community.
A key element of the \astropy Project is the core package \astropypkg, which serves as the
foundation for more specialized projects and packages.
In this article, we summarize key features in the core package as of the recent major
release, version 5.0, and provide major updates for the project.
We then discuss supporting a broader ecosystem of inter-operable packages,
including connections with several astronomical observatories and missions.
We also revisit the future outlook of the \astropy Project and the current
status of Learn Astropy.
We conclude by raising and discussing the current and future challenges facing the project.
\end{abstract}

@@

The \python programming language is a high-level, interpreted (as opposed to
compiled) programming language that has become an industry standard across many
computational domains, technological sectors, and fields of research.
This recent and rapid adoption of \python stems from the fact that it enables
scalable, time- and energy-efficient code execution \citep[e.g.,][]{Augier:2021}
with a focus on code readability, ease of use, and interoperability with other
languages.
Over the last decade, \python has grown enormously in popularity to become a
dominant programming language also in the astronomical and broader scientific
communities.
For example, Figure~\ref{fig:python-mentions} shows the number of yearly
full-text mentions of \python as compared to a few other programming languages
in refereed articles in the astronomical literature, demonstrating its nearly
exponential growth in popularity.
The adoption of \python by astronomy researchers, students, observatories, and
technical staff combined with an associated increase in awareness and interest
about open-source software tools is contributing to a paradigm shift in the way
research is done, data is analyzed, and results are shared in astronomy and
beyond.

One of the factors that has led to its rapid ascent in popularity in scientific
contexts has been the significant, volunteer-driven effort behind developing
community-oriented open-source software tools and fostering communities of users
and developers that have grown around these efforts.
Today, a broad and feature-diverse ``ecosystem'' of packages exists in the
\python scientific computing landscape: Roughly ordered from general-use to
domain-specific, this landscape now includes packages that provide core
numerical analysis functionality like \package{numpy} \citep{numpy:nature} and
\package{scipy} \citep{scipy}, visualization frameworks like
\package{matplotlib} \citep{matplotlib}, machine learning and data analysis
packages like \package{tensorflow} \citep{tensorflow}, \package{pymc3}
\citep{Salvatier:2016}, and \package{emcee} \citep{emcee}, domain-specific
libraries like \package{yt} \citep{yt:2011}, \package{plasmapy}
\citep{plasmapy}, \package{sunpy} \citep{sunpy:apj}, \package{Biopython}
\citep{biopython}, and \package{sympy} \citep{sympy} (to name a few in each
category).
The \astropypkg \citep{astropy:2013, astropy:2018} core package began in this
vein, as an effort to consolidate the development of commonly-used functionality
needed to perform astronomical research into a community-developed \python
package.

The \astropypkg core package was one of the first large, open-source \python
packages developed for astronomy and provides, among other things, software
functionality for reading and writing astronomy-specific data formats (e.g.,
FITS), transforming and representing astronomical coordinates, and representing
and propagating physical units in code.
An early description of the core functionality in \astropypkg can be found in
the first \astropy paper \citep{astropy:2013} or in detail in the core package
documentation.\footnote{\url{https://docs.astropy.org/}}
The codebase in the \astropypkg core package is now largely stable, in that the
software interface does not change without sufficient and significant
motivation, and the addition of new features into the core package has slowed
as compared to the first years of its development.
This is largely driven by the fact that the core package now represents just one
piece of the broader astronomy \python context, and much new feature development
is now happening in more specialized packages that are expanding the
capabilities of the \astropy ecosystem by building on top of the foundations
laid by the \astropypkg core package.
Because of this natural expansion, the name \astropy has grown in scope beyond a
single \python library to become ``the Astropy Project.''

@@

The \astropy Project is a community effort that represents the union of the
\astropypkg core package, the ecosystem of astronomy-specific software tools
that are interoperable with \astropypkg (Astropy Affiliated Packages),
\emph{and} the community of users, developers, and maintainers that participate
in Astropy efforts.
However, there is no institution responsible for managing the Astropy Project,
for funding or maintaining its development, or sustaining it into the future:
The Project is maintained and coordinated largely by volunteers.
While new Astropy-affiliated packages are being developed that expand upon the
core functionality in the \astropypkg package, representing a natural expansion
of the Astropy Project ecosystem, the needs of and challenges faced by the
Project are evolving.
In particular, the transition from focusing our energy on development and
maintenance of a single core package, to instead sustaining the core package and
fostering the development of the community and its expansion has been a key
issue faced by the Astropy Project in the last several years.

@@

Major versions of the core package --- that is, versions that add and/or modify
functionality --- are released approximately every six months, and are then
maintained with releases that fix issues until the next major version is
released. However, every two years a major release is designated as a long-term
support (LTS) release, which continues to be maintained for up to two years
\citep{ape2}. The motivation for LTS releases is to provide longer-term stable
versions of \astropypkg that users that require a high level of stability can
make use of if they do not always need the latest features (this could include,
for example, telescope operation or data reduction pipelines, etc.). The v5.0
release of the core package was designated as LTS, and since it was released at
the end of 2021, it will be maintained until the end of 2023.

@@

A plan for future development of \astropysubpkg{uncertainties} is to also
support error propagation, tracking covariances, for the case that uncertainties
are normally distributed rather than approximated by Monte-Carlo means. One
implementation problem to be solved is when to {\em stop} tracking covariances.
For instance, a simple operation such as subtracting the mean from $N$ data
points implies that all data points are now covariant with each other, i.e., one
has to carry $N\times N$ covariances. For a large image, that becomes not just
pointless but also prohibitive in terms of memory and CPU usage.

\paragraph{Masked quantities}  It can be useful to mask bad data. How one
supports masked data involves a number of choices, such as whether one indicates
data values are bad by a separate flag or by replacing bad data with a special
value, such as ``Not a Number'' (NaN). In \astropypkg, both approaches have
been used: a flag for masked columns in \astropyTable, mask flags and bitmaps
for N-dimensional data in \astropysubpkg{nddata}, and replacing elements with
NaN in \astropyTime. However masked quantities were poorly supported, making it
difficult to use masks in a \astropyQTable\ (in which \astropyQuantity\ is used
for all columns with units).

Unfortunately, the \texttt{MaskedArray} class from \package{numpy} only works
well with plain data arrays, hiding other metadata and attributes, for example
the \texttt{unit} of a \astropyQuantity. Consequently, a new \astropyMasked\
class was designed, based on a framework very similar to that of
\astropyDistribution, making it easy to create masked instances of other
classes. In particular, \astropyMasked\ can be used to create masked quantities
for \astropyQTable, enabling I/O of masked data from different file types.

The masked quantities work largely without any changes in higher-level objects
such as \astropySkyCoord, but work is still underway to expose the mask in those
classes, as well as to use masked arrays in \astropyTime\ instead of replacing
data with NaN.

\paragraph{Time series}  From sampling a continuous variable at fixed times to
counting events binned into time windows, many different areas of astrophysics
require the manipulation of time series. The new \astropysubpkg{timeseries}
sub-package extends the \astropyQTable\ class to support tables of data as a
function of time, where the data can either represent samples or averages over
particular time bins. The new classes offer a number of special methods to
manipulate time series (folding, resampling, etc.) and to read different data
formats (such as \mission{Kepler} light curves, etc.).

@@

Units were also integrated further within \astropypkg. For an already-defined unit-less
\astropyModel, e.g. those imported from another library, \astropymodeling can
now coerce units.
% TODO! transition words
A new module – \astropycosmologyunits – has been added to the cosmology
sub-package for defining and collecting cosmological units and equivalencies.
The unit \texttt{littleh} and equivalency
\astropyapidoc{cosmology.units.with_H0}{with\_H0} was moved from the main
\astropyunits sub-package to \astropycosmologyunits. A new unit,
\texttt{redshift}, has been added for tracking factors of cosmological redshift.
To correctly use redshift units in dimensional analysis, \texttt{redshift} is
\astropyapi{cosmology.units.dimensionless_redshift}{treated as dimensionless}
by default. To convert between redshift and other cosmological distance
measures, e.g. CMB temperature or comoving distance, the equivalency
\astropyapidoc{cosmology.units.with_redshift}{with\_redshift}
has also been added. This equivalency is actually a composite of other equivalencies,
which may be used separately.

@@

Within \astropypkg and \python there are numerous ways to represent and store
array-valued data. Some of the differences are historical:
\astropyapidoc{table.Column}{table.Column} and
\astropyapidoc{io.fits.Column}{io.fits.Column} are not the same. Some
differences are computational: \package{numpy}, \package{cupy}, and
\package{dask} arrays have almost identical APIs, but are optimized for
different use cases. Lastly, some differences are inherent: \astropyQuantity,
\astropyTime\, and \astropySkyCoord\ represent fundamentally different types of
objects.

We aim to make it possible for any array-valued data to be used as a column in a
table. A well defined protocol for mixin-columns has been developed for
\astropytable, allowing the original object to be used as a column with a
familiar API, and to ``round-trip'' through tables with no loss of data or
attributes. With this protocol, it is now possible to store \astropy native
objects -- including \astropyTime, \astropyQuantity, and \astropySkyCoord --
within a \astropyTable\ and write these to various file formats, such as FITS.
For objects not already covered by the mixin protocol, functions can be
registered with \astropyTable\ to convert any array-like object into a mixin
column. As an example, the mixin functions are used to integrate \package{dask}
arrays with \astropyTable, allowing cloud-stored or cluster-scale data to be
used as a column in a \astropyTable.

@@

The FITS standard was extended to rigorously describe time coordinates in the
World Coordinate System (WCS) framework \citep{FITS-Time:2015}. Compared to
other types of coordinates in WCS, time requires more metadata: format, scale,
position, reference, etc. This metadata had to be manually specified and the
nuances understood by the user. \astropyfits could read this data as a standard
\astropyFitsColumn\, but would not interpret the time-related metadata and
attributes. Through the support of the
\href{https://summerofcode.withgoogle.com/archive/2017/projects/4778482366152704}{Google
Summer of Code} 2017 program,\footnote{ To learn more about this project, please
see the final report
\href{https://aaryapatil.wordpress.com/2017/08/28/a-mixin-protocol-for-seamless-interoperability/}{A
mixin protocol for seamless interoperability}.}, the \astropyfits package now
interprets time data correctly, using \astropyTime\ as a `mixin' column. For
backward compatibility with manual systems, this feature may be turned off.

@@

\astropyiosubpkg{registry} is a powerful way to define input and output (I/O)
functions for \astropypkg objects, such as reading from or writing to a file,
following a given standard. \astropy uses this internally for the I/O methods in
\astropyTable\ and \astropyCosmology, and users can register custom I/O to
extend the options on these classes.

As of Astropy 5.0, the I/O registry submodule has been generalized to enable a
number of new use cases. for instance, the class-based architecture allows for
the creation of custom registries. One application is in \astropycosmology,
which has two different \textit{kinds} of registries for \astropyCosmology: one
for reading and writing files, and another for converting between python
objects. Using \astropyiosubpkg{registry} for custom classes means a developer
does not have to create a whole new I/O system and can offer to the user a
unified and familiar API.

\subsection{Precision, Accuracy, \& Reproducibility}
\label{sec:core-features-precision_accuracy_repreducibility}

@@

\paragraph{Time}
In astronomy, time accuracy down to the (nano)second is frequently important.
Whether for planning observations, crunching pulsar or VLBI data, missing
seconds meaningfully impact results. In past versions of \astropypkg, when using
\astropyTime\ in \astropytime leap seconds had to be manually applied; now, leap
seconds are applied automatically. Moreover, \astropypkg updates internal time
data-files to ensure that the correct leap second adjustment is always used. For
reproducibility, e.g., with an old code, \astropyLeapSeconds\ may also be
manually applied.

@@

Measurements of physical constants (and the units defined from them) improve
over time. Periodically the standardized systems of units and constants are
updated to reflect these improvements. For instance, in 2019 the SI system was
redefined \citep{NIST2019}, with an accompanying update to the physical
constants in
\href{https://codata.org/initiatives/data-science-and-stewardship/fundamental-physical-constants/}{\texttt{SI/CODATA
2018}}. \astropyconstants now defaults to use the \texttt{SI/CODATA 2018}
values, with the units in \astropyunits based on these constants. Most of
\astropypkg and affiliate packages build upon \astropyunits, so this update
effects the entire \astropy ecosystem.

@@

Being able to describe the position of objects, i.e., their coordinates, is
fundamental across many knowledge domains. \astropycoordinates allows one to
work with low-level positional and velocity data all the way to high-level
coordinate objects with reference frames, atmospheric information, and more. A
central feature of \astropycoordinates is the ability to transform data between
reference frames, e.g., \astropyICRS\ \citep{ICRS:1997} to \astropyGalacticLSR\
\citep{GalacticLSR:2010}. However, there have been some historical limitations
for spatially proximate transformations, impacting the usefulness of
\astropycoordinates for ground-based telescopes and astrometry within the
solar system. Transformations in the \astropyAltAz\ frame were reasonably
precise for very distant objects, but wrong by up to several arcseconds for
e.g., the location of the moon. Now these transformations are much more precise,
down to the milliarcsecond level. Similar precision improvements were made to
the
\astropyapi{coordinates.builtin_frames.HADec}{Hour Angle-Declination}
frame transformations. Additionally,
\astropyapi{coordinates.builtin_frames.BaseEclipticFrame}{Ecliptic frames}
and associated transformations have been updated to correctly reflect the “true”
and “mean” terminology.

@@

Performance in \astropypkg is addressed based on feedback and requests, and
through contributions from community members: If you find that parts of
\astropypkg are not performant, we encourage you to open a \github issue to
start a discussion with the \astropypkg core maintainers.

@@

The APE0 \citep{ape0} document lays out the principles of this governance
structure, so we refer the reader to that document for a more thorough
description, and here we only highlight some key elements. While many of these
principles were already de facto true or have been discussed organically (and
have been discussed in earlier papers in this series), the APE0-based governance
aims to provide a single place where the community can agree as a starting
point. With this in mind, it highlights the developer and user community of the
Astropy Project as the ultimate sources of authority, as well as the core
principle of ``do-ocracy'' that those who do work for the Project (be it coding,
training, or other harder-to-quantify contributions) gain more influence on the
outputs of the Project by virtue of their effort. However, APE0 adds the
concept of ``voting members'': a self-governed part of that community who are
entrusted to elect the Coordination Committee (CoCo). While the CoCo has existed
from the inception of the project, APE0 establishes a formal voting process for
CoCo members, and explicitly outlines the rights and responsibilities of the
CoCo. This role is mainly to facilitate consensus and act as the decision maker
when other mechanisms have failed, and includes powers that either
require central authority or secrets (e.g., passwords) However, APE0 also
charges the Committee to devolve responsibilities and seek community input on
these items as often as possible.

@@

\begin{figure}[t!]
    \begin{centering}
      \includegraphics[width=0.75\textwidth]{commits_committers_per_month.pdf}
        \caption{
            \textbf{Top panel:} The number of Git commits per month (defined as
            consecutive 30 day windows) merged into to the \astropypkg core
            package over the whole history of the project (up to end of 2021,
            v5.0).
            The dark line shows the mean rate using a 1 year rolling window, but
            the under-plotted gray line shows the raw monthly rate.
            The raw number clearly peaks just before feature freezes (vertical
            lines) as a flurry of activity leading up to each release, but in
            recent years the mean rate has remained steady at about $\sim$300
            commits per month.
            \textbf{Bottom panel:} Similar to above, but now showing the number
            of unique Git committers who contribute to the \astropypkg core
            package each month (again defined as consecutive 30 day windows).
            While this also shows some variation depending on proximity to
            feature freezes, in recent years the mean rate of unique
            contributors has remained consistent at about $\sim$22 committers
            per month.
        }
        \label{fig:contributor-summary:per-month}
    \end{centering}
\end{figure}

@@

The Astropy project supports the broader ecosystem by providing pre-configured
infrastructure packages that the community can use to support and maintain their
own software package infrastructure. These include tools and extensions that
enable easily generating documentation and setting up automated testing, as well
as providing package scaffolding for new projects.

@@

The Astropy Package Template helps facilitate the setup and creation of new
Python packages leveraging the Astropy ecosystem. This tool utilizes the
Cookiecutter project to walk users through the process of creating new
packages complete with documentation and testing support. Additionally, the
package template generation process includes the ability to setup
interoperability with \github, allowing for easy repository access from
documentation, as well as an example \github Actions workflow to demonstrate
the use of \github's continuous integration tooling.

@@

Several of the coordinated and affiliated packages described in
\cite{astropy:2018} have had substantial improvements. \texttt{astroquery}
\citep{astroquery} has added access to roughly a dozen new missions and data
services, including the James Webb Space Telescope (JWST) archive, and the
project has switched to a continuous release model: Every time a change is
committed to the main development branch it is published on the Python Package
Index (PyPI) and available for installation. (However, formal releases are still
done a few times per year.) \texttt{photutils} \cite {photutils} released its
first stable version, indicating that the API will change less frequently, and
there have been several significant performance improvements. \texttt{ccdproc}
\cite{ccdproc} also released a new major version, bringing better performance to
some image combination operations. The \texttt{regions} package
\cite{pyregions}, for manipulating ds9-style region definitions \cite{ds9},
added new ways to manipulate regions and introduced new region types.
\texttt{reproject}'s \cite{reproject} major new feature is a function to align
and co-add images to create a mosaic; better support for parallelization was
also added. \texttt{specutils}, the package that defines containers for 1D and
2D spectra \cite{specutils}, also had its first stable release and the addition
of classes to read JWST data. \texttt{stingray} has also had a major performance
overhaul. The package \texttt{gammapy} \cite{gammapy} also has major performance
improvements and has unified its API in preparation for its first stable
release.

@@

@@

Community Engagement

>>>>

\begin{abstract}
The \astropy Project supports and fosters the development of open-source and openly developed
\python packages that provide commonly needed functionality to the astronomical
community.
A key element of the \astropy Project is the core package \astropypkg, which serves as the
foundation for more specialized projects and packages.
In this article, we summarize key features in the core package as of the recent major
release, version 5.0, and provide major updates for the project.
We then discuss supporting a broader ecosystem of interoperable packages,
including connections with several astronomical observatories and missions.
We also revisit the future outlook of the \astropy Project and the current
status of Learn Astropy.
We conclude by raising and discussing the current and future challenges facing the project.
\end{abstract}

@@

The \python programming language is a high-level, interpreted (as opposed to
compiled) programming language that has become an industry standard across many
computational domains, technological sectors, and fields of research.
This recent and rapid adoption of \python stems from the fact that it enables
scalable, time- and energy-efficient code execution \citep[e.g.,][]{Augier:2021}
with a focus on code readability, ease of use, and interoperability with other
languages.
Over the last decade, \python has grown enormously in popularity to become a
dominant programming language, especially in the astronomical and broader scientific
communities.
For example, Figure~\ref{fig:python-mentions} shows the number of yearly
full-text mentions of \python as compared to a selection of other programming languages
in refereed articles in the astronomical literature, demonstrating its nearly
exponential growth in popularity.
The adoption of \python by astronomy researchers, students, observatories, and
technical staff combined with an associated increase in awareness and interest
about open-source software tools is contributing to a paradigm shift in the way
research is done, data is analyzed, and results are shared in astronomy and
beyond.

One of the factors that has led to its rapid ascent in popularity in scientific
contexts has been the significant, volunteer-driven effort behind developing
community-oriented open-source software tools and fostering communities of users
and developers that have grown around these efforts.
Today, a broad and feature-diverse ``ecosystem'' of packages exists in the
\python scientific computing landscape: roughly ordered from general-use to
domain-specific, this landscape now includes packages that provide core
numerical analysis functionality like \package{numpy} \citep{numpy:nature} and
\package{scipy} \citep{scipy}, visualization frameworks like
\package{matplotlib} \citep{matplotlib}, machine learning and data analysis
packages like \package{tensorflow} \citep{tensorflow}, \package{pymc3}
\citep{Salvatier:2016}, and \package{emcee} \citep{emcee}, domain-specific
libraries like \package{yt} \citep{yt:2011}, \package{plasmapy}
\citep{plasmapy}, \package{sunpy} \citep{sunpy:apj}, \package{Biopython}
\citep{biopython}, and \package{sympy} \citep{sympy} (to name a few in each
category).
The \astropypkg \citep{astropy:2013, astropy:2018} core package began in this
vein, as an effort to consolidate the development of commonly used functionality
needed to perform astronomical research into a community-developed \python
package.

The \astropypkg core package was one of the first large, open-source \python
packages developed for astronomy and provides, among other things, software
functionality for reading and writing astronomy-specific data formats (e.g.,
FITS), transforming and representing astronomical coordinates, and representing
and propagating physical units in code.
An early description of the core functionality in \astropypkg can be found in
the first \astropy paper \citep{astropy:2013} or in detail in the core package
documentation.\footnote{\url{https://docs.astropy.org/}}
The codebase in the \astropypkg core package is now largely stable, in that the
software interface does not change without sufficient and significant
motivation, and the addition of new features into the core package has slowed
as compared to the first years of its development.
This is largely driven by the fact that the core package now represents just one
piece of the broader astronomy \python context, and new feature development
is now happening in more specialized packages that are expanding the
capabilities of the \astropy ecosystem by building on top of the foundations
laid by the \astropypkg core package.
Because of this natural expansion, the name \astropy has grown in scope beyond a
single \python library to become ``the Astropy Project.''

@@

The \astropy Project is a community effort that represents the union of the
\astropypkg core package, the ecosystem of astronomy-specific software tools
that are interoperable with \astropypkg (Astropy Coordinated and Affiliated Packages),
\emph{and} the community of users, developers, and maintainers that participate
in Astropy efforts.
However, there is no single institution responsible for managing the Astropy Project,
for funding or maintaining its development, or sustaining it into the future:
The Project is maintained and coordinated largely by volunteers and contributors
who are primarily paid for other reasons (i.e., Astropy work is just a means to an end).
While new Astropy-affiliated packages are being developed that expand upon the
core functionality in the \astropypkg package, representing a natural expansion
of the Astropy Project ecosystem, the needs of and challenges faced by the
Project are evolving.
In particular, the transition from focusing our energy on development and
maintenance of a single core package, to instead sustaining the core package and
fostering the development of the community and its expansion has been a key
issue faced by the Astropy Project in the last several years.

@@

Major versions of the core package --- that is, versions that add and/or modify
functionality --- are released approximately every six months, and are then
maintained with releases that fix issues until the next major version is
released. However, every two years a major release is designated as a long-term
support (LTS) release, which continues to be maintained for up to two years
\citep{ape2}. The motivation for LTS releases is to provide longer-term stable
versions of \astropypkg that users who require a high level of stability can
make use of if they do not always need the latest features (this includes but
not limited to telescope operations and data reduction pipelines). The v5.0
release of the core package, replacing v4.0, was designated as LTS at the end of 2021;
it will be maintained until the end of 2023.

@@

A plan for future development of \astropysubpkg{uncertainties} is to
support error propagation: tracking covariances for the case that uncertainties
are normally distributed rather than approximated by Monte-Carlo means. One
implementation problem to be solved is when to {\em stop} tracking covariances.
For instance, a simple operation such as subtracting the mean from $N$ data
points implies that all data points are now covariant with each other, i.e., one
has to carry $N\times N$ covariances. For a large image, that becomes unnecessary
and computationally expensive.

\paragraph{Masked quantities}  It can be useful to mask bad data, either by marking
bad data values with a flag or by replacing them with a special
value, e.g., ``Not a Number'' (NaN). In \astropypkg, both approaches are
used: a flag for masked columns in \astropyTable, mask flags and bitmaps
for N-dimensional data in \astropysubpkg{nddata}, and replacing elements with
NaN in \astropyTime. However masked \astropyQuantity support is currently very limited,
hindering usage of masks in a \astropyQTable\ (in which \astropyQuantity\ is used
for all columns with units).

Furthermore, the \texttt{MaskedArray} class from \package{numpy} only works
well with plain data arrays, hiding other metadata and attributes, for example
the \texttt{unit} of a \astropyQuantity. Consequently, a new \astropyMasked\
class was designed, based on a framework very similar to that of
\astropyDistribution, making it easy to create masked instances of other
classes. In particular, \astropyMasked\ can be used to create masked quantities
for \astropyQTable, enabling I/O of masked data from different file types.

The masked quantities work largely without any changes in higher-level objects
such as \astropySkyCoord, but more work is needed to better integrate it with
the rest of \astropypkg, e.g., using masked arrays in \astropyTime\ instead of NaN.

\paragraph{Time series}  From sampling a continuous variable at fixed times to
counting events binned into time windows, many different areas of astrophysics
require the manipulation of time series. The new \astropysubpkg{timeseries}
sub-package extends the \astropyQTable\ class to support tables of data as a
function of time, where the data can either represent samples or averages over
particular time bins. The new classes offer a number of special methods to
manipulate time series (e.g., folding and resampling) and to read different data
formats (e.g., \mission{Kepler} light curves).

@@

Units were also integrated further within \astropypkg. For an already-defined
unit-less \astropyModel, e.g., those imported from another library,
\astropymodeling can now coerce units. A new module – \astropycosmologyunits –
has been added to the cosmology sub-package for defining and collecting
cosmological units and equivalencies. \astropycosmologyunits has a unit and
equivalencies for \texttt{littleh}, and a new unit, \texttt{redshift}, for
factors of cosmological redshift. \texttt{redshift} has a number of
equivalencies for converting between cosmological distance measures, e.g., CMB
temperature or comoving distance.

@@

Within \astropypkg and \python, there are numerous ways to represent and store
array-valued data. Some of the differences are historical:
\astropyapidoc{table.Column}{table.Column} and
\astropyapidoc{io.fits.Column}{io.fits.Column} are not the same. Some
differences are computational: \package{numpy}, \package{cupy}, and
\package{dask} arrays have almost identical APIs, but are optimized for
different use cases. Lastly, some differences are inherent: \astropyQuantity,
\astropyTime, and \astropySkyCoord\ represent fundamentally different types of
objects.

We aim to make it possible for any array-valued data to be used as a column in a
table. A well defined protocol for mixin-columns has been developed for
\astropytable, allowing the original object to be used as a column with a
familiar API, and to ``round-trip'' through tables with no loss of data or
attributes. With this protocol, it is now possible to store \astropy native
objects -- including \astropyTime, \astropyQuantity, and \astropySkyCoord\ --
within a \astropyTable\ and write these to various file formats, such as FITS.
For objects not already covered by the mixin protocol, functions can be
registered with \astropyTable\ to convert any array-like object into a mixin
column. As an example, the mixin functions are used to integrate \package{dask}
arrays with \astropyTable, allowing cloud-stored or cluster-scale data to be
used as a column in a \astropyTable.

@@

The FITS standard was extended to rigorously describe time coordinates in the
WCS framework \citep{FITS-Time:2015}. Compared to
other types of coordinates in WCS, time requires more metadata: format, scale,
position, reference, etc. This metadata had to be manually specified and the
nuances understood by the user. \astropyfits could read this data as a standard
\astropyFitsColumn\, but would not interpret the time-related metadata and
attributes. Through the support of the
\href{https://summerofcode.withgoogle.com/archive/2017/projects/4778482366152704}{Google
Summer of Code} 2017 program,\footnote{ To learn more about this project, please
see the final report
\href{https://aaryapatil.wordpress.com/2017/08/28/a-mixin-protocol-for-seamless-interoperability/}{A
mixin protocol for seamless interoperability}.} the \astropyfits package now
interprets time data correctly, using \astropyTime\ as a mixin column. For
backward compatibility, this feature may be turned off.

@@

\astropyiosubpkg{registry} is a powerful way to define input and output (I/O)
functions for \astropypkg objects, such as reading from or writing to a file,
following a given standard. \astropypkg uses this internally for the I/O methods in
\astropyTable\ and \astropyCosmology. Users can also register custom I/O to
extend the options on these classes.

As of \astropypkg v5.0, the I/O registry submodule has been generalized to enable a
number of new use cases; for instance, the class-based architecture allows
streamlining of the creation of custom registries while minimizing code duplication.
One application is in \astropycosmology,
which has two different \textit{kinds} of registries for \astropyCosmology: one
for reading and writing files, and another for converting between \python objects.

\subsection{Numerical Precision, Accuracy, \& Reproducibility}
\label{sec:core-features-precision_accuracy_repreducibility}

@@

\paragraph{Time}
In astronomy, time accuracy down to the (nano)second is frequently important.
Whether for planning observations or crunching pulsar or VLBI data, missing
seconds meaningfully impact results. In past versions of \astropypkg, when using
\astropyTime\, leap seconds had to be manually applied; now, leap
seconds are applied automatically. Moreover, \astropypkg updates internal time
data files to ensure that the correct leap second adjustment is always used. For
reproducibility with older code, \astropyLeapSeconds\ may also be
manually applied.

@@

Measurements of physical constants (and the units defined from them) improve
over time. Periodically the standardized systems of units and constants are
updated to reflect these improvements. For instance, in 2019 the SI system was
redefined \citep{NIST2019}, with an accompanying update to the physical
constants in
\href{https://codata.org/initiatives/data-science-and-stewardship/fundamental-physical-constants/}{\texttt{SI/CODATA
2018}}. \astropyconstants now defaults to using the \texttt{SI/CODATA 2018}
values, with the units in \astropyunits based on these constants. Most of
\astropypkg and affiliate packages build upon \astropyunits, so this update
affects the entire \astropy ecosystem.

@@

Being able to describe the position of objects, i.e., their coordinates, is
fundamental across many knowledge domains. \astropycoordinates allows one to
work with low-level positional and velocity data all the way to high-level
coordinate objects with reference frames, atmospheric information, and more. A
central feature of \astropycoordinates is the ability to transform data between
reference frames, e.g., \astropyICRS\ \citep{ICRS:1997} to \astropyGalacticLSR\
\citep{GalacticLSR:2010}. However, there have been some historical limitations
for spatially proximate transformations, impacting the usefulness of
\astropycoordinates for ground-based telescopes and astrometry within the
solar system. Transformations in the \astropyAltAz\ frame were reasonably
precise for very distant objects, but wrong by up to several arcseconds for
e.g., the location of the moon. Now these transformations are much more precise,
down to the milliarcsecond level. Similar precision improvements were made to
the
\astropyapi{coordinates.builtin_frames.HADec}{Hour Angle-Declination}
frame transformations. Additionally,
\astropyapi{coordinates.builtin_frames.BaseEclipticFrame}{Ecliptic frames}
and associated transformations have been updated to correctly reflect the ``true''
and ``mean" terminology.

@@

Performance in \astropypkg is addressed based on feedback and requests, and
through contributions from community members: if you find that parts of
\astropypkg are not performant, we encourage you to open a \github issue to
start a discussion with the \astropypkg core maintainers.

@@

The APE0 \citep{ape0} document lays out the principles of this governance
structure, so we refer the reader to that document for a more thorough
description, and here we only highlight some key elements. While many of these
principles were already de facto true or have been discussed organically (and
have been discussed in earlier papers in this series), the APE0-based governance
aims to provide a single place where the community can agree as a starting
point. With this in mind, it highlights the developer and user community of the
Astropy Project as the ultimate sources of authority, as well as the core
principle of ``do-ocracy'': those who do work for the Project (be it coding,
training, or other harder-to-quantify contributions) gain more influence on the
outputs of the Project by virtue of their effort. However, APE0 adds the
concept of ``voting members'': a self-governed part of that community who are
entrusted to elect the Coordination Committee (CoCo). While the CoCo has existed
from the inception of the project, APE0 establishes a formal voting process for
CoCo members, and explicitly outlines the rights and responsibilities of the
CoCo. This role is mainly to facilitate consensus and act as the decision maker
when other mechanisms have failed, and includes powers that either
require central authority or secrets (e.g., passwords). However, APE0 also
charges the Committee to devolve responsibilities and seek community input on
these items as often as possible.

@@

\begin{figure}[t!]
    \begin{centering}
      \includegraphics[width=0.75\textwidth]{commits_committers_per_month.pdf}
        \caption{
            \textbf{Top panel:} The number of Git commits per month (defined as
            consecutive 30 day windows) merged into the \astropypkg core
            package over the whole history of the project (up to end of 2021,
            v5.0).
            The dark line shows the mean rate using a 1 year rolling window, but
            the under-plotted gray line shows the raw monthly rate.
            The raw number clearly peaks just before feature freezes (vertical
            lines) as a flurry of activity leading up to each release, but in
            recent years the mean rate has remained steady at about $\sim$300
            commits per month.
            \textbf{Bottom panel:} Similar to above, but now showing the number
            of unique Git committers who contribute to the \astropypkg core
            package each month (again defined as consecutive 30 day windows).
            While this also shows some variation depending on proximity to
            feature freezes, in recent years the mean rate of unique
            contributors has remained consistent at about $\sim$22 committers
            per month.
        }
        \label{fig:contributor-summary:per-month}
    \end{centering}
\end{figure}

@@

The Astropy project supports the broader ecosystem by providing pre-configured
infrastructure packages that the community can use to support and maintain their
own software package infrastructure. These include tools and extensions that
enable generating documentation and setting up automated testing, as well
as providing package scaffolding for new projects.

@@

The Astropy Package Template helps facilitate the setup and creation of new
Python packages leveraging the Astropy ecosystem. This tool utilizes the
\href{https://github.com/cookiecutter/cookiecutter}{Cookiecutter} project to
walk users through the process of creating new packages complete with
documentation and testing support. Additionally, the package template
generation process includes the ability to setup interoperability with
\github, allowing for easy repository access from documentation, as well as an
example \github Actions workflow to demonstrate the use of \github's
continuous integration tooling.

@@

Several of the coordinated and affiliated packages described in
\cite{astropy:2018} have had substantial improvements. \texttt{astroquery}
\citep{astroquery} has added access to roughly a dozen new missions and data
services, including the James Webb Space Telescope (JWST) archive, and the
project has switched to a continuous release model: every time a change is
committed to the main development branch it is published on the Python Package
Index (PyPI) and available for installation. (However, formal releases are still
done a few times per year.) \texttt{photutils} \cite{photutils} released its
first stable version, indicating that the API will change less frequently, and
there have been several significant performance improvements. \texttt{ccdproc}
\cite{ccdproc} also released a new major version, bringing better performance to
some image combination operations. The \texttt{regions} package
\cite{pyregions}, for manipulating ds9-style region definitions \cite{ds9},
added new ways to manipulate regions and introduced new region types.
\texttt{reproject}'s \cite{reproject} major new feature is a function to align
and co-add images to create a mosaic; better support for parallelization was
also added. \texttt{specutils}, the package that defines containers for 1D and
2D spectra \cite{specutils}, also had its first stable release and the addition
of classes to read JWST data. \texttt{stingray} has also had a major performance
overhaul. The package \texttt{gammapy} \cite{gammapy} also has major performance
improvements and has unified its API in preparation for its first stable
release.

@@

% \begin{itemize}
% \item {\bf User forums:} The Astropy Project has historically maintained a number of avenues for users to seek help or gain access to the developer community. This includes the Astropy mailing list, the Python in Astronomy Facebook group, and the Astropy Slack workspace. The Moore Foundation grant allows this Slack space to
% be on a paid plan; additionally NumFOCUS has negotiated a special rate for
% open-source projects. While the Slack and Astropy-dev mailing lists are primarily used to discuss the project direction and updates, it was noted that the use of a Facebook community could present a barrier to open source. The Astropy Project identified a need for a public, archived, searchable, and easily-to-navigate interface for users to ask for help Accordingly, we have commissioned
% a Discourse site which is more open than the Facebook group and more user-oriented
% than Astropy Slack. A benefit of the Moore Foundation grant is that Astropy
% developers are able to invoice as independent contractors the time they
% spend helping users on these forums.
% \end{itemize}

@@

- Astropy is established; challenges faced have evolved
-

\begin{itemize}
\item Attracting new contributors when the code has become quite complex, project established (astropy as an "institution", walled-off group). (how to attract? Fast feedback and code reviews, ...)
\item Support for existing contributors: Contributor and maintainer mentoring (...)
\item Infrastructure: Long-term / sustained funding for maintaining infrastructure (grants last 3 years; infrastructure evolves, nature of open source ecosystem)
\item As ecosystem grows, making sure users know about all of the functionality (workshops, tutorials, educational materials, discoverability, accessibility) (Continued engagement with user community)
\end{itemize}
